### 2023-03-02 calculus.tex
Add some references through section 2

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
type error localization in bidirectional systems. Namely, the calculus provides a judgmental
framework with a unified theory for the design of typing semantics that support ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
lambda calculus (GTLC), extending it with binary product types in \cref{sec:calculus-products}. All
rules and theorems may be found in the supplementary appendix, alongside a complete mechanization in
the Agda proof assistant, which is discussed briefly in \cref{sec:calculus-agda}. With products, we
then consider how error localization is complicated by patterned let expressions in
\cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
to more complex typing features, such as System F-style polymorphism.

@@

Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, establishes that, under
the typing context $\ctx$, the expression $e$ synthesizes or infers the type $\TMV$. \emph{Type
analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
where an expression of type $\TMV$ is expected. We may give these typing semantics traditionally,
without considering error localization and recovery.

@@

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. The analysis rule employs the judgment
$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type}
[citation], i.e. it may be considered an arrow type. Defined in \cref{fig:calculus-matched-arrow},
this notion is purely a technical mechanism to avoid duplication of rules related to arrow types.

@@

    \item \begin{enumerate}
        \item If there does not exist $\TMV$ such that $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, then for
          all $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
          the case that $\markless{\ECMV}$.
        \item If there does not exist $\TMV$ such that $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, then for
          all $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
          the case that $\markless{\ECMV}$.
      \end{enumerate}
  \end{enumerate}
\end{theorem}
%
That is, since error marks indicate type errors in the program, no marks should be introduced
when marking a well-typed program, a property given by the judgment $\markless{\ECMV}$. The
converse---that at least one mark is inserted into ill-typed programs---is also true.
% TODO latter is easier to do in Agda, but maybe it is weaker?

@@

The semantics and metatheory presented above, including the extension with binary products, have
been fully mechanized in the Agda proof assistant [citation]. Though the mechanization's
documentation contains more detailed discussion regarding technical decisions made therein, we
highlight some important aspects here.

>>>>

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
type error localization in bidirectional systems. Namely, the calculus provides a judgmental
framework with a unified theory for the design of typing semantics that support ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
lambda calculus (GTLC), extending it with binary product types in \cref{sec:calculus-products}. All
rules and theorems may be found in the supplementary appendix, alongside a complete mechanization in
the Agda proof assistant \cite{norell:thesis} (discussed in \cref{sec:calculus-agda}). With
products, we then consider how error localization is complicated by patterned let expressions in
\cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
to more complex typing features, such as System F-style polymorphism.

@@

Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, establishes that, under
the typing context $\ctx$, the expression $e$ synthesizes or locally infers \cite{Localinf} the type
$\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, states that the expression
$\EMV$ may appear where an expression of type $\TMV$ is expected. We may give these typing semantics
traditionally, without considering error localization and recovery.

@@

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. The analysis rule employs the judgment
$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type}
\cite{cimini:2016}, i.e. it may be considered an arrow type. Defined in
\cref{fig:calculus-matched-arrow}, this notion is purely a technical mechanism to avoid duplication
of rules related to arrow types.

@@

    \item \begin{enumerate}
        \item If there does not exist $\TMV$ such that $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, then for
          all $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
          the case that $\markless{\ECMV}$.
        \item If there does not exist $\TMV$ such that $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, then for
          all $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
          the case that $\markless{\ECMV}$.
      \end{enumerate}
  \end{enumerate}
\end{theorem}
%
That is, since error marks indicate type errors in the program, no marks should be introduced
when marking a well-typed program, a property given by the judgment $\markless{\ECMV}$. The
converse---that at least one mark is inserted into ill-typed programs---is also true.

@@

The semantics and metatheory presented above, including the extension with binary products, have
been fully mechanized in the Agda proof assistant. Though the mechanization's documentation contains
more detailed discussion regarding technical decisions made therein, we highlight some important
aspects here.
